---
title: "Model Fitting"
author: "Christopher Conley, Pei Wang, Jie Peng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```


Fitting the spacemap model to 1662 genomic copy number alternations (CNA) and 1595 protein expressions across 77 heterogeneous breast cancer tumor samples requires a process that guards against overfitting. We used *CV.Vote* for tuning parameter selection and *Boot.Vote* for building an ensemble network that finds real biological signal amidst millions of possible interactions in the data set. This model fitting process demands a powerful computational framework that leverages parallelism. 


*Note:* The following code is for llustration purposes only and it is not recommended to evaluate the CV.Vote (see [cvVote](https://topherconley.github.io/spacemap/reference/cvVote.html)) or Boot.Vote (see [bootEnsemble](https://topherconley.github.io/spacemap/reference/bootEnsemble.html) and [bootVote](https://topherconley.github.io/spacemap/reference/bootVote.html)) steps on a simple laptop computer^[Actual model fitting employed 124 workers on a computer cluster. The last tuning grid iteration took `cvVote()` 34 hours.]. 

## Setup

Load the `ExpressionSet` data objects containing protein expressions and genomic copy number, which was generated in [the previous step](https://topherconley.github.io/neta-bcpls/data.html).

```{r}
suppressPackageStartupMessages(library(Biobase))
cnaset <- readRDS("data/cna-expression-set.rds") 
cna <- t(Biobase::exprs(cnaset))
protset <- readRDS("data/prot-expression-set.rds")
pexp <- t(Biobase::exprs(protset))
#standardize
pexp <- scale(pexp); cna <- scale(cna);
```

Load the cross validation test sets, which have balanced molecular subtypes across the test sets.

```{r}
testSetIds <- readRDS(file = "data/prot_cv_10k_test_sets77.rds")
#create training sets.
allSampleIds <- seq_len(nrow(pexp))
trainSetIds <- lapply(testSetIds, function(testSetId) setdiff(allSampleIds, testSetId))
```

Load the last iteration of the tuning grid used in the CV.Vote step. 

```{r}
tmap <- readRDS(file = "data/prot_last_tune_grid.rds")
```

Establish a parallel backend to evaluate model fits efficiently.

```{r}
library(doParallel)
library(parallel)
ncores <- detectCores()  - 1
cl <- makeCluster(ncores)
registerDoParallel(cl)
```

Create a directory to save all the model fits from `cvVote`. 

```{r}
rp <- "~/scratch-data/neta-bcpls/mfits/"
system(paste("mkdir -p ", rp))
```


## CV.Vote

The CV.Vote step will estimate optimal tuning parameters for the BCPLS application, which are denoted as $\hat \lambda^*_1, \hat  \lambda^*_2, \hat \lambda^*_3$. 


```{r}
library(spacemap)
spmapcv <- cvVote(Y = pexp, X = cna, 
                  trainIds = trainSetIds, testIds = testSetIds, 
                  method = "spacemap", tuneGrid = tmap,
                  resPath = rp, aszero = 1e-4,
                  tol = 1e-4, cd_iter = 60e7)
```

The estimated optimal tuning parameters (previously computed) are: 

```{r, eval=TRUE, echo = FALSE}
readRDS(file = "data/best_prot_tune_params.rds")
```

## Boot.Vote

Fit the spaceMap model on bootstrapped data replicates subject to $\hat \lambda^*_1, \hat  \lambda^*_2, \hat \lambda^*_3$, which will create a bootstrap ensemble of networks. Then, construct a final network through majority vote on the bootstrap ensemble of networks.  

```{r}
ens <- bootEnsemble(Y = pexp, X = cna, tune = spmapcv$minTune, method = "spacemap", B = 1000, 
                    aszero = 1e-4, tol = 1e-4, seed = 55139L)
ensbv <- bootVote(ens)
```


## Next Step in Analysis

Please see the [Annotation](https://topherconley.github.io/neta-bcpls/annot.html) article for the next step in the analysis. 



