---
title: "Network Analysis"
author: "Christopher Conley, Pei Wang, Jie Peng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(message=F, warning=F, cache=TRUE)
```

We take the opportunity to illustrate the analysis of the CNA-mRNA network (i.e. `RNA-net`) in this document since the the CNA-protein network (i.e. `prot-net`) was used for illustration of the network analysis toolkit in the spacemap R package's [vignette](https://topherconley.github.io/spacemap/articles/neta.html). Please refer to the vignette prior to inspecting this analysis. The two documents share virtually identical analytical steps, but this document differs in the response data and  we don't go to great lengths to explain the input arguments to the toolkit.

## Input

Load the gene coordinate annotations for mRNA expressions and genomic CNA.

```{r}
library(Biobase)
yinfo <- pData(featureData(readRDS(file = "data/rna-expression-set.rds")))
xinfo <- pData(featureData(readRDS(file = "data/cna-expression-set.rds")))
```

Load the *Boot.Vote* CNA-mRNA network. 

```{r}
net <- readRDS(file = "model-fits/smap_rna_boot_vote.rds")
names(net) <- c("yy", "xy")
rownames(net$xy) <- xinfo$id; colnames(net$xy) <- yinfo$id;
rownames(net$yy) <- yinfo$id; colnames(net$yy) <- yinfo$id;
```

Load the degree distributions for the ensemble of bootstrapped networks. 

```{r}
bdeg <- readRDS(file = "model-fits/smap_rna_boot_degree.rds")
names(bdeg) <- c("xy", "yy")
colnames(bdeg$xy) <- xinfo$id; colnames(bdeg$yy) <- yinfo$id;
```

Load the Gene Ontology mappings for enrichment analysis. 

```{r}
go2eg <- readRDS("data/rna-go-bp-to-entrez-gene-list.rds")
library(AnnotationDbi)
#human readable 
process_alias <- Term(names(go2eg))
```


## Map Annotations

Convert the Boot.Vote network into an *igraph* object and map the annotations onto the network. 

```{r}
library(spacemap)
ig <- spacemap::adj2igraph(yy = net$yy, xy = net$xy, yinfo = yinfo, xinfo = xinfo)
```

If we query the attribute names of the nodes in the graph, we notice that the columns of `xinfo` and `yinfo` have been applied. 

```{r}
vertex_attr_names(graph = ig)
```

The igraph package has a number of ways to access the annotation information. For example, if we wish to confirm the chromosome location of GRB7, we can easily query: 

```{r}
vertex_attr(graph = ig, name = "chr", index = V(ig)[alias %in% "GRB7"])
```

## Hub Analysis
We first prioritize the CNA- and mRNA- hubs. If  the `bdeg` argument is specified, then we rank the hubs according to the average degree rank. Accordingly, the most highly ranked hubs will have the most consistently high degree across network ensemble. 

To rank the mRNA nodes, use the `rankHub` command and simply specify the `level = "y"` argument. 

```{r}
ig <- rankHub(ig = ig, bdeg = bdeg$yy, level = "y")
```

To rank the CNA nodes, specify the `level = "x"` argument. 

```{r}
ig <- rankHub(ig = ig, bdeg = bdeg$xy, level = "x")
```

Alternatively, if the `bdeg` argument is not available, the ranking is according to degree in the final network (coded by "ig"). 

```{r}
tmp <- rankHub(ig = ig,  level = "x")
```

### Identify cis and trans
Next label $x-y$ edges as being regulated in cis or in trans. The *GenomicRanges* package and the genomic coordinates `chr,start,end` columns of `xinfo` and `yinfo` are required for this step. If you don't have the package  [GenomicRanges](http://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) installed yet, try: 

Now we can label the $x-y$ edges with either "cis" or "trans" in the `cis_trans` edge attribute of `ig`.

```{r}
ig <- cisTrans(ig = ig, level = "x-y", cw = 2e6)
```

### Report 
We then report a well-organized table, as seen in Table 3 of the spaceMap publication. The `top` argument allows us to control how many hubs are displayed.

```{r}
xhubs <- reportHubs(ig, top = 7, level = "x")
```

```{r, echo=FALSE}
kable(xhubs, row.names = FALSE)
```

Similarly, we can report the top 10 mRNA hubs,  their degrees in the  final network, and a description of each hub, if the `description` column was specified in `yinfo`. 

```{r}
yhubs <- reportHubs(ig, top = 10, level = "y")
```

```{r, echo=FALSE}
kable(yhubs, row.names = FALSE)
```

### GO-neighbor percentage

A _CNA neighborhood_ comprises all mRNA nodes that are directly connected to a CNA hub by an edge. CNA neighborhoods  represent direct perturbations to the proteome by amplifications or deletions in the DNA. To quantify their functional relevance, we compute a score called the _GO-neighbor percentage_. Two mRNA nodes are called GO-neighbors if they share a common GO term in the same CNA neighborhood. We postulate that a high percentage of GO-neighbors within a CNA neighborhood associates the CNA hub with more functional meaning. These scores, as presented in Figure 5 of the publication, can be generated with a GO mapping to the mRNAs as follows. 

```{r}
hgp <- xHubEnrich(ig = ig, go2eg = go2eg)
```

```{r, echo=FALSE}
kable(hgp, row.names = FALSE)
```

## Module Analysis

There are many criteria to define modules of a network. This toolkit  allows users to import the module membership information by themselves (see `mods` argument in [modEnrich](https://topherconley.github.io/spacemap/reference/modEnrich.html)).  

In the spaceMap publication, we use the edge-betweenness algorithm in *igraph*.

```{r}
library(igraph)
mods <- cluster_edge_betweenness(ig)
```

The main goal of module analysis is identifying modules that are functionally enriched. 
The `modEnrich` function will test for significantly over-represented GO-terms (or any other valid functional grouping) within a module using hyper-geometric tests.  


In  the current example, only the mRNA nodes have functional mapping and we specify this through the `levels = "y"` argument. If both predictors and responses have functional mapping in the `go2eg` argument, then we can specify `levels = c("y","x")`.  Other arguments are available to control the enrichment testing; see [modEnrich](https://topherconley.github.io/spacemap/reference/modEnrich.html) for more details. 

```{r}
outmod <- modEnrich(ig = ig, mods = mods, levels = "y", go2eg = go2eg, process_alias = process_alias, prefix = "R")
```

The output of the module analysis is a list of 3 objects.

```{r}
names(outmod)
```

+  `ig` is the input igraph network object updated with a "process_id" attribute for nodes affiliated with a significant GO-term. The "process_id" and "module" attributes together are
useful for visualizing which nodes are enriched for a specific biological function. 

+ `etab` is the polished module enrichment table to report significant GO terms, the representation of the GO term in the module relative to the size of the GO term, and which CNA hubs  belong to the module. The top ten hits as in Table S.9 of the spaceMap publication's supplementary materials are as follows: 


```{r, eval = FALSE}
outmod$etab[1:10,]
```


```{r, echo=FALSE}
knitr::kable(outmod$etab[1:10,], row.names = FALSE)
```

+ `eraw` contains details for each (module, GO-term) pair that was subjected to the hyper-geometric test. This output gives the user more control by reporting all tests, the relative over-representation of a GO-term in that module, the raw P-value, and the adjusted P-value. 

```{r, eval = FALSE}
outmod$eraw[1:5,]
```

```{r, echo=FALSE}
knitr::kable(outmod$eraw[1:5,], row.names = FALSE)
```

## Export for Visualization

There are many tools for network visualization. In the publication of spaceMap, we exported the annotated *igraph* network object `ig` to the "graphml" format, which maintained all the attributes associated with nodes when read into *Cytoscape* for visualization. 

```{r}
igraph::write_graph(graph = outmod$ig, file = "neta/spacemap-rna-boot-vote.graphml", format = "graphml")
```

Here we list all the attributes associated with the nodes that can be used in tandem with Cytoscape's filtering functions to identify specific nodes of interest. 

```{r}
vertex_attr_names(outmod$ig)
```

We describe some of the most useful attributes for visualization:

+ 'name': the unique node ID 
+ 'alias': the node alias (e.g. gene symbol ERBB2)
+ 'chr,start,end,strand': the gene coordinates of nodes
+ 'description': any note (e.g. breast cancer oncogene)
+ 'levels': indicates whether the node belongs to predictors "x" or responses "y"
+ 'rank_hub': the rank of the hub within its level (e.g. a value of "1" for a node of level "x" means that it is the most consistently high degree $x$ node in the network. ) 
+ 'regulates_in_cis': list of genes regulated in cis 
+ 'module': the module ID that the node belongs to. 
+ 'process_id': the significant GO-term(s) associated with the node. 

Also the edge attributes are exported to 'graphml' format.

```{r}
edge_attr_names(outmod$ig)
```

+ 'levels' indicates whether an edge is $x-y$ or $y-y$.
+ 'cis_trans' indicates whether an edge is regulated in cis or in trans. 

## Summary

This toolkit is useful for analyzing and summarizing the output of  the spaceMap model fit in the context of integrative genomics. The biological context mapped onto the network object can easily be exported to  existing network visualization tools like *Cytoscape*. 


