---
title: "Spacemap Protein-Network Analysis Notebook"
output:
  html_document: default
  html_notebook: default
---

### Feature attribute loading

Load the proteogenomics data. 

```{r}
suppressPackageStartupMessages(library(Biobase))
rnaset <- readRDS(file = "~/repos/neta-bccptac/data/rna-expression-set.rds")
protset <- readRDS(file = "~/repos/neta-bccptac/data/prot-expression-set.rds")
cnaset <- readRDS(file = "~/repos/neta-bccptac/data/cna-expression-set.rds")
```

Combine the feature attributes between the response (i.e. protein) and the predictors (i.e. CNA). 

```{r}
yy_node_attributes <- featureData(protset)
yy_node_attributes <- pData(yy_node_attributes)
yy_node_attributes <- yy_node_attributes[,-10]
xy_node_attributes <- featureData(cnaset)
xy_node_attributes <- pData(xy_node_attributes)
xy_yy_node_attributes <- rbind(xy_node_attributes, yy_node_attributes)
rownames(xy_yy_node_attributes) <- NULL
```

Load the network analysis functions. 

```{r}
suppressMessages(library(igraph))
source("~/repos/bioc_spacemap/R/networkAnalysis.R")
```

Import the Boot.vote fit as well as the degree distributions of the bootstrap replicates. 

```{r}
prot_spmap <- readRDS(file = "~/scratch-data/cptac/rep/spacemap_prot_boot_vote.rds")
ig<-  adj2igraph(YY = prot_spmap$bv$ParCor, 
                 XY = prot_spmap$bv$Gamma, 
                 info = xy_yy_node_attributes)
igy <- adj2igraph(YY = prot_spmap$bv$ParCor, info = yy_node_attributes, dropnull = F)
#degree distribution of bootstrap replicates
protbdeg <- readRDS(file = "~/scratch-data/cptac/rep/spacemap_prot_boot_degree.rds")
xbdeg <- protbdeg$xbdeg
ybdeg <- protbdeg$ybdeg
```


### Discovery of -cis and -trans regulation 


```{r}
ig <- setEdgeTypeAttr(ig, "x", "y")
#E(ig)[edge_type %in% "x--y"]
ig <- setEdgeTypeAttr(ig, "y", "y")
#E(ig)[edge_type %in% "y--y"]
ig <- setCisTransAttr(ig = ig, et = c("x--y", "y--y"))
#E(ig)$cis_trans
```


```{r}
vseqx <- V(ig)[type %in% "x"]
xdct <- setCisTransDistr(ig = ig, vseq = vseqx)
ig <- xdct$ig
dct <- xdct$dct
```


### Inspect which CNA are most stable

```{r}
pc <- potentialCis(pact = as_ids(V(ig)[type %in% "x"]),
                   preg = xy_yy_node_attributes$id[xy_yy_node_attributes$type == "y"], 
                   info = xy_yy_node_attributes,
                   YY = prot_spmap$bv$ParCor, XY = prot_spmap$bv$Gamma, 
                   cw = 2e6, is = T)
```


```{r}
colnames(xbdeg) <- featureNames(cnaset)
xhubtab <- renderXhubTable(ig = ig, dct = dct, pc = pc, bdeg = xbdeg)
suppressMessages(library(xlsx))
```

```{r}
netadir <- "c:/Users/topher/Shared/scratch-data/neta-bccptac/neta/spacemap/prot/"
library(xtable)
#Part of table  in section 4.3
cap <- "Prioritized CNA hubs of \texttt{spaceMap prot-net} in each row with the number of -cis and -trans edges emanating from CNA hub, potential number of -cis relationships for the Hub, and which genes are in -cis with the CNA hub."
print(xtable(xhubtab[,c(1,5:7)], display  = c("d","s", "s", "d", "s"), caption = cap),
      include.rownames=FALSE,
      type = "latex", 
      file = file.path(netadir, "spacemap-cna-hub-with-prot.tex"))
#excel file
xlsx::write.xlsx(x = xhubtab, file = file.path(netadir, "spacemap-xhub-prot.xlsx"))
#supplementary section 
capsup <- "Prioritized CNA hubs of \texttt{spaceMap prot-net} in each row with gene coordinates, the number of -cis and -trans edges emanating from the CNA hub, potential number of -cis relationships for the Hub, which genes are in -cis with the CNA hub, and the mean (std. deviation) rank of CNA degree across 1000 bootstrap replications."
print(xtable(xhubtab[,c(1,3:8)], display  = c("d","s", "d", "d","s", "d", "s", "s"), caption = capsup),
      include.rownames=FALSE,
      type = "latex", 
      file = file.path(netadir, "spacemap-cna-hub-with-prot-supp.tex"))
```

## GO Enrichment

Load the Protein GO universe. 

```{r}
go2eg <- readRDS(file = "~/repos/neta-bccptac/data/prot-go-bp-to-entrez-gene-list.rds")
eg2go <- readRDS(file = "~/repos/neta-bccptac/data/prot-go-entrez-gene-to-bp-list.rds")
```


Obtain network modules using the `Girvan-Newman` algorithm for clust

```{r}
cebout <- cebModules(ig, glb = 15, prefix = "P")
ig <- cebout$ig
lmods <- cebout$module_list
yy_lmods <- cebout$yy_module_list
ceb <- cebout$community_obj
cross_edges <- crossing(communities = ceb, graph = ig)
cross_edges2 <- names(cross_edges)[cross_edges]
cross_edge_tab <- splitEdgeVector(cross_edges2, ig)
```

Report the proportion of edges which have both vertices in the same GO category. 

```{r}
total_edges <- ecount(igy)
iyy_edges <- as_ids(E(igy))  
edge_pair <- strsplit(iyy_edges, split = "|", fixed = TRUE)
ngopairs <- sapply(edge_pair, function(ep) any(sapply(go2eg, function(egs) all(ep %in% egs))))
prop_go_pairs <- sum(ngopairs) / total_edges
sum(ngopairs)
prop_go_pairs
```

Identify the GO pair proportion for each CNA hub by considering any two response nodes with a common CNA hub as adjacent.

```{r}
hgp <- hubGOproportion(ig, go2eg)
xlsx::write.xlsx(x = hgp, file = file.path(netadir, "spacemap-prot-cna-neighborhood-go-modularity.xlsx"), row.names = F)
saveRDS(object = hgp, file = file.path(netadir, "spacemap-prot-hub-neighborhood-go-modularity-obs.rds"))
obs_go_prop <- mean(hgp[,"Proportion"])
obs_go_prop
```

Evaluate the mean GO pair proportion under random graphs with fixed degree distribution. 

```{r}
xy <- prot_spmap$bv$Gamma + 0
yy <- prot_spmap$bv$ParCor + 0
# set.seed(58759)
# if(file.exists(file.path(netadir, "go_100_prop.rds"))) { 
#   rnd_go_prop <- readRDS(file.path(netadir, "go_100_prop.rds"))
# } else { 
#   rnd_go_prop <- replicate(n = 100, rndHubGoProportion(xy, yy, info = xy_yy_node_attributes, go2eg = go2eg))
#   saveRDS(list(rnd = rnd_go_prop, obs = obs_go_prop), file = file.path(netadir, "go_100_prop.rds"))
#   mean(rnd_go_prop)
#   sd(rnd_go_prop)
# }
```


```{r}
set.seed(58759)
if(file.exists(file.path(netadir, "prot_net_rhgps_100.rds"))) {
  rhgps <- readRDS(file.path(netadir, "prot_net_rhgps_100.rds"))
} else {
  rigs <-lapply(1:50, function(i) rndCGGraph(xy = xy, yy = yy, info =  xy_yy_node_attributes))
  saveRDS(rigs, file = file.path(netadir, "prot_net_rnd_graphs_50.rds"))
  lrhgps <- lapply(rigs, function(rig) { 
    rhgp <- hubGOproportion(rig, go2eg)
  })
  rhgps <- do.call(what = rbind, args = lrhgps)
  saveRDS(rhgps, file = file.path(netadir, "prot_net_rhgps_50.rds"))
  colMeans(rhgps[,c("m1","m2","m3")])
  apply(as.matrix(rhgps[,c("m1","m2","m3")]), 2, sd)
}
```


Evaluate which GO terms enrich the protein-network modules. 

```{r}
modenrich <- moduleEnrichment(lmods = yy_lmods, go2eg = go2eg, igy = igy, fdr_thresh = 0.05)
head(modenrich)
```


The P-value distribution is well behaved. 

```{r}
library(ggplot2)
qplot(modenrich$pvalue) + xlab("P-value of Module Enrichment Analysis")
```


Format the module enrichment results for displaying in a table.

```{r}
minGO <- 5
modenrichtab <- renderModuleTable(ig, lmods, modenrich, minGO = minGO) 
cap <- "GO-enriched modules (edge-betweeness) in spacemap protein-network."
print(xtable(modenrichtab, display = c("d", "s", "s", "s", "s"), 
             caption = cap, digits = -2),
      include.rownames = FALSE, 
      type = "latex", 
      file = file.path(netadir, "spacemap-mod-enrichment-prot.tex"))
```

```{r}
ig <- setGOTags(lmods = lmods, modenrich = modenrich, 
                ig = ig, go2eg = go2eg, info = xy_yy_node_attributes)
```

```{r}
lmod_attr <- lapply(names(lmods), getModuleInfo,  
                    ig = ig,
                    go2eg = go2eg,
                    lmods = lmods,
                    modenrich = modenrich,
                    info = xy_yy_node_attributes, 
                    cross_edge_tab = cross_edge_tab, 
                    minGO = minGO)
names(lmod_attr) <- names(lmods)
  
library(xlsx)
nonull <- lapply(seq_along(lmod_attr), function(i) { 
  write.xlsx(x = lmod_attr[[i]], 
             file = paste0(netadir, "spacemap-modules-prot.xlsx"), 
             sheetName = names(lmod_attr)[i], 
             row.names = FALSE, append = TRUE)})
```

Save igraph object.

```{r}
write.graph(ig, file = file.path(netadir, "spacemap-prot-boot-vote.gml"), format = "gml")
write.graph(ig, file = file.path(netadir, "spacemap-prot-boot-vote.graphml"), format = "graphml")
write.graph(ig, file = file.path(netadir, "spacemap-prot-boot-vote.el"), format = "edgelist")
```

Write modules with enriched terms to a gml file. 

```{r}
for(mod_name in names(lmod_attr)) { 
  module_ids <- lmod_attr[[mod_name]]$Unique_ID
  mod_subgraph <- induced_subgraph(graph = ig, vids = V(ig)[name %in% module_ids])
  write.graph(mod_subgraph, file = paste0(netadir, "spacemap", mod_name, "_mod_bv_prot.graphml"), format = "graphml")  
}
```


```{r}
save.image(file = file.path(netadir, "spacemap-neta-prot.RData"))
```


