---
title: "Preparing Gene Ontology universes"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

A Gene Ontology (GO) universe is composed of all the biological processess specifically found among the protein and mRNA features that were used to learn `prot-net` and `RNA-net`. This document shows how the GO universes were constructed for GO enrichment analysis of genes within network modules of  `prot-net` and `RNA-net`.

## Protein GO universe

Load the protein features and map refseq to entrez genes. 

```{r,eval=FALSE}
protset <- readRDS(file = "../data/prot-expression-set.rds")
suppressMessages(library(mygene))
p2r77map <- queryMany(qterms = featureNames(protset), scopes = "refseq",
                      fields = c("entrezgene"))
p2r77map2 <- data.frame(refseq = p2r77map@listData$query, entrezgene = p2r77map@listData$entrezgene)
saveRDS(p2r77map2, file = "../data/prot-refseq-to-entrezgene-map.rds")
```

Read in the previous code block's results. 

```{r}
p2r77map2 <- readRDS(file = "../data/prot-refseq-to-entrezgene-map.rds")
```

Get the entrez genes with GO mappings. 

```{r}
suppressMessages(library(org.Hs.eg.db))
x <- org.Hs.egGO
hs_feature_list <- function(db, eg) { 
  mk <- mappedkeys(db)
  xx <- as.list(db[eg[eg %in% mk]])
}
# Get the entrez gene identifiers that are mapped to a GO ID
mapped_genes <- mappedkeys(x)
entrezgene.qe.go <- as.character(p2r77map2$entrezgene[ p2r77map2$entrezgene %in% mapped_genes])
leg2go <-  hs_feature_list(db = org.Hs.egGO, eg = entrezgene.qe.go) 
```

Map the entrez genes to GO biological process terms with some valid evidence (i.e. not "ND").

```{r}
library(foreach)
leg2go_filtered <- foreach(got = leg2go, eg = names(leg2go)) %do% { 
  bp_domain <- sapply(got, function(x) x[["Ontology"]]) == "BP"
  #bp_domain <- TRUE
  not_nd_domain <- sapply(got, function(x) x[["Evidence"]]) != "ND"
  bp_got <- got[bp_domain & not_nd_domain]
  sapply(bp_got, function(x) x[["GOID"]]) 
}
names(leg2go_filtered) <- names(leg2go)
leg2go_filtered <- leg2go_filtered[sapply(leg2go_filtered, length) != 0]
```

Reverse the mapping to GO -> entrez gene and keep those mappings that have at least 15 but no more than 300 GO terms. 

```{r}
suppressMessages(library(topGO))
go2eg <- inverseList(leg2go_filtered)
#switch to refseq id because of graph id and to include all isoforms
go2eg <- lapply(go2eg, function(x) as.character(p2r77map2$refseq[ p2r77map2$entrezgene %in% x ]))
#remove duplicate entrez genes
go2eg <- lapply(go2eg, function(x) x[!duplicated(x)])
#saveRDS(go2eg, file = "~/repos/neta-bccptac/data/no-len-trim-prot-go-bp-to-entrez-gene-list.rds")
gosize <- sapply(go2eg, length)
go2eg <- go2eg[ gosize >= 15 & gosize <= 300]
#assure each go term has a list of non-duplicate ids
stopifnot(all(sapply(go2eg, anyDuplicated) == 0))
eg2go <- inverseList(go2eg)
```


Save the protein GO universe.

```{r, eval = FALSE}
saveRDS(go2eg, file = "../data/prot-go-bp-to-entrez-gene-list.rds")
saveRDS(eg2go, file = "../data/prot-go-entrez-gene-to-bp-list.rds")
```

Clear the workspace for the RNA analysis. 

```{r}
rm(list = ls())
```

## RNA GO universe

Get the entrez genes with GO mappings. 

```{r}
rnaset <- readRDS(file = "../data/rna-expression-set.rds")
rna_entrez_ids <- featureNames(rnaset)
library(IRanges)
library(org.Hs.eg.db)
hs_feature_list <- function(db, eg) { 
  mk <- mappedkeys(db)
  xx <- as.list(db[eg[eg %in% mk]])
}
leg2go <- hs_feature_list(db = org.Hs.egGO, eg = rna_entrez_ids)
```


Map the entrez genes to GO BP terms with some valid evidence (not "ND").

```{r}
library(foreach)
leg2go_filtered <- foreach(got = leg2go, eg = names(leg2go)) %do% { 
  bp_domain <- sapply(got, function(x) x[["Ontology"]]) == "BP"
  #bp_domain <- TRUE
  not_nd_domain <- sapply(got, function(x) x[["Evidence"]]) != "ND"
  bp_got <- got[bp_domain & not_nd_domain]
  sapply(bp_got, function(x) x[["GOID"]]) 
}
names(leg2go_filtered) <- names(leg2go)
leg2go_filtered <- leg2go_filtered[sapply(leg2go_filtered, length) != 0]
```

Reverse the mapping to GO -> entrez gene and keep those mappings that have at least 15 but no more than 300 GO terms. 

```{r}
suppressPackageStartupMessages(library(topGO))
go2eg <- inverseList(leg2go_filtered)
#remove duplicate entrez genes
go2eg <- lapply(go2eg, function(x) x[!duplicated(x)])
saveRDS(go2eg, file = "~/repos/neta-bccptac/data/no-len-trim-rna-go-bp-to-entrez-gene-list.rds")
gosize <- sapply(go2eg, length)
go2eg <- go2eg[ gosize >= 15 & gosize <= 300]
#number of go terms satisfying conditions
length(go2eg)
#assure each go term has a list of non-duplicate ids
stopifnot(all(sapply(go2eg, anyDuplicated) == 0))
eg2go <- inverseList(go2eg)
```

Save the RNA GO universe. 

```{r, eval = FALSE}
saveRDS(go2eg, file = "../data/rna-go-bp-to-entrez-gene-list.rds")
saveRDS(eg2go, file = "../data/rna-go-entrez-gene-to-bp-list.rds")
```


### Next Step in Analysis

Please see the [Network Analysis](https://topherconley.github.io/neta-bcpls/neta/neta-bcpls.html) article for the next step in the analysis. 

#### Session Info

```{r}
sessionInfo()
```

